---
title: "NOAA Storm Database Analysis"
author: "Derwin McGeary"
date: "August 21, 2015"
output: html_document
---

# This is the title which I have yet to write

## Questions

Your data analysis must address the following questions:

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
1. Across the United States, which types of events have the greatest economic consequences?

Consider writing your report as if it were to be read by a government or municipal manager who might be responsible for preparing for severe weather events and will need to prioritize resources for different types of events. However, there is no need to make any specific recommendations in your report.

## Synopsis

## Data processing

```{r, cache=TRUE}
stormData <- read.csv("repdata_data_StormData.csv.bz2")
```

## Analysis

```{r}

```

## Results


You can also embed plots, for example:

```{r}

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
