---
title: "NOAA Storm Database Analysis"
author: "Derwin McGeary"
date: "August 21, 2015"
output: html_document
---

# This is the title which I have yet to write

## Questions

Your data analysis must address the following questions:

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most *harmful* with respect to *population health*?
1. Across the United States, which types of events have the *greatest economic consequences*?

Consider writing your report as if it were to be read by a government or municipal manager who might be responsible for preparing for severe weather events and will need to prioritize resources for different types of events. However, there is no need to make any specific recommendations in your report.

## Synopsis

## Data processing

We will be interested in the columns EVTYPE, CROPDMG, CROPDMGEXP, PROPDMG, PROPDMGEXP, INJURIES, and FATALITIES. So we can read in the data and check how many NA values are in those columns.

```{r, cache=TRUE}
stormData <- read.csv("repdata_data_StormData.csv.bz2")
sum(is.na(stormData$CROPDMG))
sum(is.na(stormData$CROPDMGEXP))
sum(is.na(stormData$PROPDMG))
sum(is.na(stormData$PROPDMGEXP))
sum(is.na(stormData$INJURIES))
sum(is.na(stormData$FATALITIES))

```

We are interested in events which caused damage, injury, and loss of life, so we can discard any rows where these are all zero.

```{r}
stormData <- subset(stormData, INJURIES + FATALITIES + PROPDMG + CROPDMG > 0)
```
We may also look at a histogram of events by date.
```{r}
hist(as.Date(stormData$BGN_DATE,"%m/%d/%Y"), breaks=62)
```

There is a large disparity between data collected before and after around 1995. As the earlier data is much smaller in volume and in quality, we may discard it.

```{r}
stormData <- subset(stormData, as.Date(stormData$BGN_DATE, "%m/%d/%Y")>as.Date("01/01/1995","%m/%d/%Y"))
```

### Calculating total damage amounts

As listed in the Storm Data Preparation guidance, there are multipliers in columns PROPDMGEXP and CROPDMGEXP, where K, M, and B correspond to thousands, millions, and billions respectively.

```{r}
summary(stormData$PROPDMGEXP[stormData$PROPDMG > 0])
summary(stormData$CROPDMGEXP[stormData$CROPDMG > 0])
```

As there are relatively few exceptions, with K and M dominating in each case, we will ignore the others rather than guessing (i.e. assume a multiplicative factor of 1 for the CROPDMG/PROPDMG numbers).

```{r}
exp2mult <- function(x) switch(x, B=1000000000, M=1000000, K=1000, 1)
stormData$PROPDMGMULT <- sapply(toupper(as.character(stormData$PROPDMGEXP)), exp2mult)
stormData$CROPDMGMULT <- sapply(toupper(as.character(stormData$CROPDMGEXP)), exp2mult)
stormData$CROPDMGTOTAL <- stormData$CROPDMG * stormData$CROPDMGMULT
stormData$PROPDMGTOTAL <- stormData$PROPDMG * stormData$PROPDMGMULT
```

### Event Types
As listed in [the Storm Data Preparation guidance, on page 6](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf), there are 48 Event Names, corresponding to the column EVTYPE in our data. We'll put them in a variable, `eventnames`.
```{r}
eventnames <- c("Astronomical Low Tide", "Avalanche", "Blizzard", "Coastal Flood", "Cold/Wind Chill", "Debris Flow", "Dense Fog", "Dense Smoke", "Drought", "Dust Devil", "Dust Storm", "Excessive Heat", "Extreme Cold/Wind Chill", "Flash Flood", "Flood", "Frost/Freeze", "Funnel Cloud", "Freezing Fog", "Hail", "Heat", "Heavy Rain", "Heavy Snow", "High Surf", "High Wind", "Hurricane (Typhoon)", "Ice Storm", "Lake-Effect Snow", "Lakeshore Flood", "Lightning", "Marine Hail", "Marine High Wind", "Marine Strong Wind", "Marine Thunderstorm Wind", "Rip Current", "Seiche", "Sleet", "Storm Surge/Tide", "Strong Wind", "Thunderstorm Wind", "Tornado", "Tropical Depression", "Tropical Storm", "Tsunami", "Volcanic Ash", "Waterspout", "Wildfire", "Winter Storm", "Winter Weather")
EVENTNAMES <- gsub("\\ ","",toupper(eventnames))
```
Now, let's look at the real data.
```{r}
print(evtypes<-length(unique(stormData$EVTYPE)))
```
Unfortunately there are `r evtypes` different names in EVTYPE. We shall try to make this more manageable.

```{r}
stormData$EVTYPE <- toupper(stormData$EVTYPE)
stormData$EVTYPE <- gsub("\\ ","",stormData$EVTYPE)
subb <- function(x,y) stormData$EVTYPE <- gsub(x,y,stormData$EVTYPE)
stormData$EVTYPE <- gsub("SEVERE","",stormData$EVTYPE)
stormData$EVTYPE <- gsub("HVY","HEAVY",stormData$EVTYPE)
stormData$EVTYPE <- gsub(".*RAINFALL.*","HEAVYRAIN",stormData$EVTYPE)
stormData$EVTYPE <- gsub(".*LIG[HN]T[N]*ING.*","LIGHTNING",stormData$EVTYPE)
stormData$EVTYPE <- gsub("TSTM", "THUNDERSTORMWIND", stormData$EVTYPE)
stormData$EVTYPE <- gsub("T[H]*U[N]*[D]*[E]*[R]*","THUNDER",stormData$EVTYPE)
stormData$EVTYPE <- gsub("STROM","STORM",stormData$EVTYPE)
stormData$EVTYPE <- gsub("^THUNDERSTORM.*","THUNDERSTORMWIND",stormData$EVTYPE)
stormData$EVTYPE <- gsub(".*FLASHFLOOD.*","flashflood",stormData$EVTYPE)
stormData$EVTYPE <- gsub(".*FLOOD.*","FLOOD",stormData$EVTYPE)
stormData$EVTYPE <- gsub("flashflood","FLASHFLOOD",stormData$EVTYPE)
stormData$EVTYPE <- gsub("^HAIL.*","HAIL",stormData$EVTYPE)
stormData$EVTYPE <- gsub("^TROPICALSTORM.*","STORM",stormData$EVTYPE)
stormData$EVTYPE <- gsub("^HURRICANE.*","HURRICANE(TYPHOON)",stormData$EVTYPE)
stormData$EVTYPE <-gsub("^TYPHOON$","HURRICANE(TYPHOON)", stormData$EVTYPE)
stormData$EVTYPE <- gsub("^TORNADO.*","TORNADO",stormData$EVTYPE)
stormData$EVTYPE <- gsub(".*SURGE.*","STORMSURGE/TIDE",stormData$EVTYPE)
stormData$EVTYPE <- gsub("HEAVYSHOWER","HEAVYRAIN",stormData$EVTYPE)
stormData$EVTYPE <- gsub("HEAVYSNOWSHOWER","HEAVYSNOW",stormData$EVTYPE)
stormData$EVTYPE <-gsub(".*AVALAN.*", "AVALANCHE", stormData$EVTYPE)
stormData$EVTYPE <-gsub(".*BLIZZARD.*","BLIZZARD", stormData$EVTYPE)
stormData$EVTYPE <-gsub(".*WINT.*WEA.*","WINTERWEATHER", stormData$EVTYPE)
stormData$EVTYPE <-gsub("WINTRY.*","WINTERWEATHER", stormData$EVTYPE)
stormData$EVTYPE <-gsub(".*WINTERSTORM.*","WINTERSTORM", stormData$EVTYPE)
stormData$EVTYPE <-gsub(".*CURRENT.*","RIPCURRENT", stormData$EVTYPE)
stormData$EVTYPE <-gsub(".*SURF.*","HIGHSURF", stormData$EVTYPE)
stormData$EVTYPE <- gsub("^COLD$","COLD/WINDCHILL", stormData$EVTYPE)
stormData$EVTYPE <- gsub("^COLDWAVE$","COLD/WINDCHILL", stormData$EVTYPE)

stormData$EVTYPE <-gsub(".*EXTRE
                        MECOLD.*","EXTREMECOLD/WINDCHILL", stormData$EVTYPE)
stormData$EVTYPE <-gsub("EXTREMEWINDCHILL","EXTREMECOLD/WINDCHILL", stormData$EVTYPE)
stormData$EVTYPE <-gsub("EXTREMEHEAT","HEAT", stormData$EVTYPE)
stormData$EVTYPE <- gsub("^HEATWAVE$","HEAT", stormData$EVTYPE)

stormData$EVTYPE <- gsub(".*FREEZE.*","FROST/FREEZE", stormData$EVTYPE)
stormData$EVTYPE <- gsub(".*FREEZINGRAIN.*","SLEET", stormData$EVTYPE)
stormData$EVTYPE <- gsub("FREEZINGDRIZZLE","SLEET", stormData$EVTYPE)
stormData$EVTYPE <- gsub("STORMFORCEWINDS","HIGHWIND", stormData$EVTYPE)
stormData$EVTYPE <- gsub(".*WATERSPOUT.*","WATERSPOUT", stormData$EVTYPE)
stormData$EVTYPE <- gsub(".*FIRE.*","WILDFIRE", stormData$EVTYPE)
stormData$EVTYPE <- gsub(".*DROUGHT.*","DROUGHT", stormData$EVTYPE)
stormData$EVTYPE <- gsub("WINDS","WIND", stormData$EVTYPE)
stormData$EVTYPE <- gsub("^WIND$", "STRONGWIND", stormData$EVTYPE) # STRONGWIND is the lower category of STRONG and HIGH
stormData$EVTYPE <- gsub("^HIGHWIND.*","HIGHWIND", stormData$EVTYPE)
stormData$EVTYPE <- gsub("WHIRLWIND","TORNADO", stormData$EVTYPE)
stormData$EVTYPE <- gsub("WINDWIND","WIND", stormData$EVTYPE)
stormData$EVTYPE <- gsub(".*FROST.*","FROST/FREEZE", stormData$EVTYPE)
stormData$EVTYPE <- gsub("SMALLHAIL","HAIL", stormData$EVTYPE)
stormData$EVTYPE <- gsub("^STORM$","THUNDERSTORMWIND", stormData$EVTYPE)
stormData$EVTYPE <- gsub("^RAIN$","HEAVYRAIN", stormData$EVTYPE)
stormData$EVTYPE <- gsub("^RECORDHEAT$","EXCESSIVEHEAT", stormData$EVTYPE)
stormData$EVTYPE <- gsub("^RECORDCOLD$","EXTREMECOLD/WINDCHILL", stormData$EVTYPE)

print(evtypes<-length(unique(stormData$EVTYPE)))
```

We have corrected for the multifarious and creative misspellings of "thunderstorm" and attempted to select the most specific category where more than one is mentioned. This is still nowhere near complete, but it covers the categories with the largest figures. The difference is under eight percent for each of the measurements we care about.

```{r}
unique(stormData$EVTYPE)
cleanData <- subset(stormData, EVTYPE %in% EVENTNAMES)
(sum(stormData$FATALITIES)-sum(cleanData$FATALITIES))/sum(stormData$FATALITIES)
(sum(stormData$INJURIES)-sum(cleanData$INJURIES))/sum(stormData$INJURIES)
(sum(stormData$PROPDMGTOTAL)-sum(cleanData$PROPDMGTOTAL))/sum(stormData$PROPDMGTOTAL)
(sum(stormData$CROPDMGTOTAL)-sum(cleanData$CROPDMGTOTAL))/sum(stormData$CROPDMGTOTAL)
stormData <- cleanData
```

## Analysis

So we can aggregate the data by our two public health columns, INJURIES and FATALITIES, and our two economic columns, PROPDMG and CROPDMG. As the latter two are both dollar amounts, we will sum them. INJURIES and FATALITIES are fundamentally different categories, so we will not combine them.

```{r}
stormEconAgg <- aggregate(cbind(PROPDMGTOTAL,CROPDMGTOTAL)~EVTYPE, data=stormData, FUN=sum)
stormEconAgg$TotalDamage <- stormEconAgg$PROPDMGTOTAL + stormEconAgg$CROPDMGTOTAL
stormEconAgg$PercentDamage <- 100 * stormEconAgg$TotalDamage/sum(stormEconAgg$TotalDamage)
stormEconAgg <- stormEconAgg[order(-stormEconAgg$TotalDamage),]
stormHealthAgg <- aggregate(cbind(INJURIES,FATALITIES)~EVTYPE, data=stormData, FUN=sum)
```

## Results


You can also embed plots, for example:

```{r}
barplot(height = stormEconAgg$TotalDamage[1:5], names.arg = stormEconAgg$EVTYPE[1:5])
stormHealthAgg <- stormHealthAgg[order(-stormHealthAgg$FATALITIES),]
barplot(height = stormHealthAgg$FATALITIES[1:5], names.arg = stormHealthAgg$EVTYPE[1:5])
```
