---
title: "Floods and Hurricanes cause more property damage, while Tornados, Heat, and Floods have the greatest impact on human health"
author: "Derwin McGeary"
date: "August 21, 2015"
output: html_document
---

## Questions

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most *harmful* with respect to *population health*?
1. Across the United States, which types of events have the *greatest economic consequences*?

Consider writing your report as if it were to be read by a government or municipal manager who might be responsible for preparing for severe weather events and will need to prioritize resources for different types of events. However, there is no need to make any specific recommendations in your report.

## Synopsis

We have analysed the data from the NOAA Storm Database in order to rank the health and economic effects of weather events. Economic damage was assessed by adding the damage to crops and the damage to property. Economically, floods and hurricanes are the most destructive events by a large margin.

Public health impact was assessed by considering the number of injuries and deaths reported with each event. No effort has been made to combine these factors. Tornados, excessive heat, and heat are the most lethal events. The greatest number of injuries is caused by tornados too, followed by floods, excessive heat, and thunderstorms.

## Data processing

Our data was retrieved from https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2 on August 20th 2015. We will be interested in the columns `EVTYPE`, `CROPDMG`, `CROPDMGEXP`, `PROPDMG`, `PROPDMGEXP`, `INJURIES`, and `FATALITIES`. So we can read in the data and check how complete those columns are.

```{r, cache=TRUE}
stormData <- read.csv("repdata_data_StormData.csv.bz2")
sum(is.na(stormData[23:28]))
```

We are interested in events which caused damage, injury, and loss of life, so we can discard any rows where these are all zero.

```{r}
stormData <- subset(stormData, INJURIES + FATALITIES + PROPDMG + CROPDMG > 0)
```

We may also look at a histogram of events by date.

```{r}
hist(as.Date(stormData$BGN_DATE,"%m/%d/%Y"), breaks="years", main="Storm Database Events by Year", xlab="Year")
```

There is a large disparity between data collected before and after around 1995. As the earlier data is much smaller in volume and lower in quality, we may discard it.

```{r}
stormData <- subset(stormData, as.Date(stormData$BGN_DATE, "%m/%d/%Y")>as.Date("01/01/1995","%m/%d/%Y"))
```

### Calculating total damage amounts

As listed in the [Storm Data Preparation guidance, on page 12](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf), there are multipliers in the columns PROPDMGEXP and CROPDMGEXP, where K, M, and B correspond to thousands, millions, and billions respectively. Let us compare this with the actual dataset.

```{r}
summary(stormData$PROPDMGEXP[stormData$PROPDMG > 0])
summary(stormData$CROPDMGEXP[stormData$CROPDMG > 0])
```

There are other values in there too. As there are relatively few other values, with K and M dominating in each case, we will ignore the others rather than guessing (i.e. assume a multiplicative factor of 1 for the CROPDMG/PROPDMG numbers).

```{r}
exp2mult <- function(x) switch(x, B=1000000000, M=1000000, K=1000, 1)
stormData$PROPDMGMULT <- sapply(toupper(as.character(stormData$PROPDMGEXP)), exp2mult)
stormData$CROPDMGMULT <- sapply(toupper(as.character(stormData$CROPDMGEXP)), exp2mult)
stormData$CROPDMGTOTAL <- stormData$CROPDMG * stormData$CROPDMGMULT
stormData$PROPDMGTOTAL <- stormData$PROPDMG * stormData$PROPDMGMULT
```

### Event Types

As listed in [the Storm Data Preparation guidance, on page 6](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf), there are 48 Event Names, corresponding to the column EVTYPE in our data. We'll put them in a variable, `eventnames`.
```{r}
eventnames <- c("Astronomical Low Tide", "Avalanche", "Blizzard", "Coastal Flood", "Cold/Wind Chill", "Debris Flow", "Dense Fog", "Dense Smoke", "Drought", "Dust Devil", "Dust Storm", "Excessive Heat", "Extreme Cold/Wind Chill", "Flash Flood", "Flood", "Frost/Freeze", "Funnel Cloud", "Freezing Fog", "Hail", "Heat", "Heavy Rain", "Heavy Snow", "High Surf", "High Wind", "Hurricane (Typhoon)", "Ice Storm", "Lake-Effect Snow", "Lakeshore Flood", "Lightning", "Marine Hail", "Marine High Wind", "Marine Strong Wind", "Marine Thunderstorm Wind", "Rip Current", "Seiche", "Sleet", "Storm Surge/Tide", "Strong Wind", "Thunderstorm Wind", "Tornado", "Tropical Depression", "Tropical Storm", "Tsunami", "Volcanic Ash", "Waterspout", "Wildfire", "Winter Storm", "Winter Weather")
EVENTNAMES <- gsub("\\ ","",toupper(eventnames))
```

Now, let's look at the real data.

```{r}
print(evtypes<-length(unique(stormData$EVTYPE)))
```

Unfortunately there are `r evtypes` different names in EVTYPE. We shall try to make this more manageable.

```{r}
stormData$EVTYPE <- toupper(stormData$EVTYPE)
stormData$EVTYPE <- gsub("\\ ","",stormData$EVTYPE)
subb <- function(x,y) stormData$EVTYPE <- gsub(x,y,stormData$EVTYPE)
stormData$EVTYPE <- gsub("SEVERE","",stormData$EVTYPE)
stormData$EVTYPE <- gsub("HVY","HEAVY",stormData$EVTYPE)
stormData$EVTYPE <- gsub(".*RAINFALL.*","HEAVYRAIN",stormData$EVTYPE)
stormData$EVTYPE <- gsub(".*LIG[HN]T[N]*ING.*","LIGHTNING",stormData$EVTYPE)
stormData$EVTYPE <- gsub("TSTM", "THUNDERSTORMWIND", stormData$EVTYPE)
stormData$EVTYPE <- gsub("T[H]*U[N]*[D]*[E]*[R]*","THUNDER",stormData$EVTYPE)
stormData$EVTYPE <- gsub("STROM","STORM",stormData$EVTYPE)
stormData$EVTYPE <- gsub("^THUNDERSTORM.*","THUNDERSTORMWIND",stormData$EVTYPE)
stormData$EVTYPE <- gsub(".*FLASHFLOOD.*","flashflood",stormData$EVTYPE)
stormData$EVTYPE <- gsub(".*FLOOD.*","FLOOD",stormData$EVTYPE)
stormData$EVTYPE <- gsub("flashflood","FLASHFLOOD",stormData$EVTYPE)
stormData$EVTYPE <- gsub("^HAIL.*","HAIL",stormData$EVTYPE)
stormData$EVTYPE <- gsub("^TROPICALSTORM.*","STORM",stormData$EVTYPE)
stormData$EVTYPE <- gsub("^HURRICANE.*","HURRICANE(TYPHOON)",stormData$EVTYPE)
stormData$EVTYPE <-gsub("^TYPHOON$","HURRICANE(TYPHOON)", stormData$EVTYPE)
stormData$EVTYPE <- gsub("^TORNADO.*","TORNADO",stormData$EVTYPE)
stormData$EVTYPE <- gsub(".*SURGE.*","STORMSURGE/TIDE",stormData$EVTYPE)
stormData$EVTYPE <- gsub("HEAVYSHOWER","HEAVYRAIN",stormData$EVTYPE)
stormData$EVTYPE <- gsub("HEAVYSNOWSHOWER","HEAVYSNOW",stormData$EVTYPE)
stormData$EVTYPE <-gsub(".*AVALAN.*", "AVALANCHE", stormData$EVTYPE)
stormData$EVTYPE <-gsub(".*BLIZZARD.*","BLIZZARD", stormData$EVTYPE)
stormData$EVTYPE <-gsub(".*WINT.*WEA.*","WINTERWEATHER", stormData$EVTYPE)
stormData$EVTYPE <-gsub("WINTRY.*","WINTERWEATHER", stormData$EVTYPE)
stormData$EVTYPE <-gsub(".*WINTERSTORM.*","WINTERSTORM", stormData$EVTYPE)
stormData$EVTYPE <-gsub(".*CURRENT.*","RIPCURRENT", stormData$EVTYPE)
stormData$EVTYPE <-gsub(".*SURF.*","HIGHSURF", stormData$EVTYPE)
stormData$EVTYPE <- gsub("^COLD$","COLD/WINDCHILL", stormData$EVTYPE)
stormData$EVTYPE <- gsub("^COLDWAVE$","COLD/WINDCHILL", stormData$EVTYPE)

stormData$EVTYPE <-gsub(".*EXTRE
                        MECOLD.*","EXTREMECOLD/WINDCHILL", stormData$EVTYPE)
stormData$EVTYPE <-gsub("EXTREMEWINDCHILL","EXTREMECOLD/WINDCHILL", stormData$EVTYPE)
stormData$EVTYPE <-gsub("EXTREMEHEAT","HEAT", stormData$EVTYPE)
stormData$EVTYPE <- gsub("^HEATWAVE$","HEAT", stormData$EVTYPE)

stormData$EVTYPE <- gsub(".*FREEZE.*","FROST/FREEZE", stormData$EVTYPE)
stormData$EVTYPE <- gsub(".*FREEZINGRAIN.*","SLEET", stormData$EVTYPE)
stormData$EVTYPE <- gsub("FREEZINGDRIZZLE","SLEET", stormData$EVTYPE)
stormData$EVTYPE <- gsub("STORMFORCEWINDS","HIGHWIND", stormData$EVTYPE)
stormData$EVTYPE <- gsub(".*WATERSPOUT.*","WATERSPOUT", stormData$EVTYPE)
stormData$EVTYPE <- gsub(".*FIRE.*","WILDFIRE", stormData$EVTYPE)
stormData$EVTYPE <- gsub(".*DROUGHT.*","DROUGHT", stormData$EVTYPE)
stormData$EVTYPE <- gsub("WINDS","WIND", stormData$EVTYPE)
stormData$EVTYPE <- gsub("^WIND$", "STRONGWIND", stormData$EVTYPE) # STRONGWIND is the lower category of STRONG and HIGH
stormData$EVTYPE <- gsub("^HIGHWIND.*","HIGHWIND", stormData$EVTYPE)
stormData$EVTYPE <- gsub("WHIRLWIND","TORNADO", stormData$EVTYPE)
stormData$EVTYPE <- gsub("WINDWIND","WIND", stormData$EVTYPE)
stormData$EVTYPE <- gsub(".*FROST.*","FROST/FREEZE", stormData$EVTYPE)
stormData$EVTYPE <- gsub("SMALLHAIL","HAIL", stormData$EVTYPE)
stormData$EVTYPE <- gsub("^STORM$","THUNDERSTORMWIND", stormData$EVTYPE)
stormData$EVTYPE <- gsub("^RAIN$","HEAVYRAIN", stormData$EVTYPE)
stormData$EVTYPE <- gsub("^RECORDHEAT$","EXCESSIVEHEAT", stormData$EVTYPE)
stormData$EVTYPE <- gsub("^RECORDCOLD$","EXTREMECOLD/WINDCHILL", stormData$EVTYPE)

print(evtypes<-length(unique(stormData$EVTYPE)))
```

We have corrected for the multifarious and creative misspellings of "thunderstorm" and attempted to select the most specific category where more than one is mentioned. This is still nowhere near complete, but it covers the categories with the largest figures. The difference is under five percent for each of the measurements we care about.

```{r}
cleanData <- subset(stormData, EVTYPE %in% EVENTNAMES)
(sum(stormData$FATALITIES)-sum(cleanData$FATALITIES))/sum(stormData$FATALITIES)
(sum(stormData$INJURIES)-sum(cleanData$INJURIES))/sum(stormData$INJURIES)
(sum(stormData$PROPDMGTOTAL)-sum(cleanData$PROPDMGTOTAL))/sum(stormData$PROPDMGTOTAL)
(sum(stormData$CROPDMGTOTAL)-sum(cleanData$CROPDMGTOTAL))/sum(stormData$CROPDMGTOTAL)
stormData <- cleanData
stormData$EVTYPE <- eventnames[match(stormData$EVTYPE, EVENTNAMES)]
```

## Analysis

So we can aggregate the data by our two public health columns, INJURIES and FATALITIES, and our two economic columns, PROPDMG and CROPDMG. As the latter two are both dollar amounts, we will sum them. INJURIES and FATALITIES are fundamentally different categories, so we will not combine them (an injury is not half a death).

```{r}
stormEconAgg <- aggregate(cbind(PROPDMGTOTAL,CROPDMGTOTAL)~EVTYPE, data=stormData, FUN=sum)
stormEconAgg$TotalDamage <- stormEconAgg$PROPDMGTOTAL + stormEconAgg$CROPDMGTOTAL
stormEconAgg$PercentDamage <- 100 * stormEconAgg$TotalDamage/sum(stormEconAgg$TotalDamage)
stormEconAgg <- stormEconAgg[order(-stormEconAgg$TotalDamage),]
stormHealthAgg <- aggregate(cbind(INJURIES,FATALITIES)~EVTYPE, data=stormData, FUN=sum)
# Making this long form for ease of plotting in one panel
strmHealth <- reshape(stormHealthAgg, varying = c("FATALITIES","INJURIES"), v.names="Number",timevar = "Type", times = c("Fatalities","Injuries"), direction = "long")
```

## Results

### Economic Impact
Now we can view the economic data, ordered by severity.

```{r}
library(ggplot2)
ggplot(stormEconAgg, aes(x=reorder(EVTYPE,TotalDamage), y=TotalDamage/1000000)) + labs(list(title = "Economic Damage due to Storm Events", x="Event Type", y="Million USD")) + theme_grey(base_size = 8) + geom_bar(stat='identity') + coord_flip()
```

Clearly floods and hurricanes are the most destructive events by a large margin. Indeed, combined they cause `r sum(stormEconAgg$PercentDamage[1:2])` percent of total economic damage in this dataset.

### Health Impact
```{r}
ggplot(strmHealth, aes(x=reorder(EVTYPE,Number), y=Number)) + labs(list(title = "Public Health Effects of Storm Events", x="Event Type", y= "Number")) + theme_grey(base_size = 8) +  geom_bar(stat='identity') + facet_grid(. ~ Type) + coord_flip()
```

Tornados, excessive heat, and simply heat are the most lethal events. The greatest number of injuries is caused by tornados too, followed by floods, excessive heat, and thunderstorms.

## Further research
It would be useful to have a more granular analysis, at the very least, separating agricultural economic damage out. Also, there is a list stating locations of injury or death, which would also enable a more focused response, and possible intervention.